Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-05 paper/paper.tex
more compact illustrations
less detail, contributions upfront

\subsection*{Contributions}

@@

A probabilistic program, or query, mostly runs deterministic
code, except for certain checkpoints, in which probabilities are
involved, and normal, linear execution of the program is
disrupted. In Anglican and similar languages there are two types
of such checkpoints:

@@

Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

Because of the checkpoints, Anglican programs must allow the
inference algorithm to step in, recording information and
affecting control flow. This can be implemented through
coroutines/cooperative multitasking, and parallel execution/preemptive
multitasking, as well as through explicit
maintenance of program continuations at checkpoints. Anglican
follows the latter option. Clojure is
a functional language, and continuation-passing style (CPS)
transformation is a well-developed technique in the area of
functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher burden
on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. The Anglican `compiler', represented by a set of
functions in the \texttt{anglican.trap} namespace, accepts a
Clojure subset and transforms it into a variant of CPS
representation, which allows inference algorithms to intervene
in the execution flow at probabilistic checkpoints.

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm}
%and \texttt{fm}
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

Operational semantics of Anglican queries is different from that
of Clojure code, therefore queries must be called through
inference algorithms, rather than `directly'.  The
\texttt{anglican.inference} namespace declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
This function accepts an Anglican query and
returns a lazy sequence of weighted samples from the
distribution defined by the query.  When inference
is performed on an Anglican query, the query is run by a
particular inference algorithm. Inference algorithms must
provide an implementation for \texttt{infer}, as well as
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

Finally, Anglican queries use `primitive', or commonly known
and used, distributions, to draw random samples and condition
observations. Many primitive distributions are provided by the
\texttt{anglican.runtime} namespace, and an additional
distribution can be defined by the user by implementing a
particular set of functions for the distribution
(via Clojure's protocol mechanism). The \texttt{defdist} macro
provides a convenient syntax for defining primitive distributions.

@@

Having defined the query, we are now ready to run the query
using an inference algorithm. Function \texttt{doquery} provided
by the \texttt{anglican.core} namespace accepts the inference
algorithm, the query, and optional parameters, and returns a
lazy sequence of samples. We use here the inference algorithm
called Lightweight Metropolis-Hastings (LMH). The algorithm is
somewhat slow to converge but can be used with any Anglican
query, and should be robust enough for our simple problem.

@@

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
    \caption{Arrival time distribution for a single customer.}
    \label{fig:dist-same}
\end{figure}

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
    \caption{Arrival time distributions for two customers.}
    \label{fig:dist-diff}
\end{figure}

@@

The compilation relies on the Clojure \textit{macro} facility,
and implemented as a library of \textit{functions} in namespace
\texttt{anglican.trap}, which are invoked by macros.  The CPS
transformation is organized in top-down manner.  The top-level
function is  \texttt{cps-of-expression}, which receives an
expression and a continuation, and returns the expression in the
CPS form. For example, the CPS transformation of constant
\texttt{1} with continuation \texttt{cont} thus takes the
following form:
\begin{lstlisting}[style=default]
=> (cps-of-expression 1 'cont)
(cont 1 $state)
\end{lstlisting}

@@

In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

@@

In the implementation of Anglican, inference algorithms are
instantiations of the (ad-hoc) polymorphic function
\texttt{infer} declared as Clojure's multimethod in the
\texttt{anglican.{\linebreak[0]}inference} namespace. The function accepts an
algorithm identifier, a query --- the probabilistic program in
which to perform the inference, an initial value for the query,
and optional algorithm parameters.

@@

An execution of the query may result in the following
sequence of checkpoint invocations:
\begin{lstlisting}[style=default]
    (sample 'C1 ...)
    (sample 'C2 ...)
    (sample 'C2 ...)
    (sample 'C1 ...)
    (sample 'C1 ...)
    (sample 'C1 ...)
    (sample 'C2 ...)
    (sample 'C3 ...)
\end{lstlisting}
According to the addressing scheme, the addresses generated
for this invocations are
\begin{lstlisting}[style=default]
    [C1 0]
    [C2 0]
    [C2 1]
    [C1 16]
    [C1 17]
    [C1 18]
    [C2 16]
    [C3 0]
\end{lstlisting}
If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
\begin{lstlisting}[style=default]
    (sample 'C1 ...)
    (sample 'C2 ...)
    (sample 'C1 ...)
    (sample 'C1 ...)
    (sample 'C2 ...)
    (sample 'C2 ...)
    (sample 'C3 ...)
\end{lstlisting}
The addresses for the new sequence are
\begin{lstlisting}[style=default]
    [C1 0]
    [C2 0]
    [C1 16]
    [C1 17]
    [C2 16]
    [C2 17]
    [C3 0]
\end{lstlisting}
Despite the change, the correspondence between checkpoints of
similar types occurring in similar positions (relative positions
in contiguous subsequences of a certain type) is preserved, and
drawn values can be reused efficiently:

@@

This way, each occurrence of a checkpoint has unique address,
but small disturbances --- removal or addition of a single or just
a few checkpoints --- are unlikely to derail the entire sequence.
The probability of derailment depends on the padding. The
padding can be safely, and without any impact on performance,
set to rather large numbers. However, rounding up to a multiple
of 16 proved to be appropriate for all practical purposes.

Function \texttt{checkpoint-id} in the
\texttt{anglican.inference} namespace automates generation of
checkpoint addresses and can be used from any inference
algorithm.

>>>>

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
	\item Design and implementation of a probabilistic
		programming language Anglican involving tight bilateral
		integration with a general-purpose programming language.
	\item Definition of semantics of probabilistic forms in
		Anglican which facilitates full separation of
		probabilistic programs from inference algorithms.
	\item Stateless representation of exchangeable
		random procedures, or processes, seamlessly integrated
		into pure functional computation.
	\item Techniques for efficient and compact implementation of
		inference algorithms, such as representation of
		inference results as a lazy sequence of samples and 
		a novel scheme for addressing of checkpoints.
\end{itemize}

@@

A probabilistic program, or query, mostly runs deterministic
code, except for certain checkpoints, in which probabilities are
involved, and normal, linear execution of the program is
disrupted.
\iftoggle{full}{%
In Anglican and similar languages there are two types
of such checkpoints:

@@

}{%
}%
Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
\iftoggle{full}{%
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

}{%
}%
Because of the checkpoints, Anglican programs must allow the
inference algorithm to step in, recording information and
affecting control flow. This can be implemented through
coroutines/cooperative multitasking, and parallel execution/preemptive
multitasking, as well as through explicit
maintenance of program continuations at checkpoints. Anglican
follows the latter option.
\iftoggle{full}{%
Clojure is
a functional language, and continuation-passing style (CPS)
transformation is a well-developed technique in the area of
functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher burden
on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. 
}{%
}%
The Anglican `compiler'\iftoggle{full}{, represented by a set of
functions in the \texttt{anglican.trap} namespace,}{} accepts a
Clojure subset and transforms it into a variant of CPS
representation, which allows inference algorithms to intervene
in the execution flow at probabilistic checkpoints.

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros\iftoggle{full}{ (defined in the
\texttt{anglican.emit} namespace)}{} which call the CPS
transformations and define Clojure values suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm} (Anglican counterpart of Clojure \texttt{defn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

Operational semantics of Anglican queries is different from that
of Clojure code, therefore queries must be called through
inference algorithms, rather than `directly'. \iftoggle{full}{The
\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
This function accepts an Anglican query and
returns a lazy sequence of weighted samples from the
distribution defined by the query.  When inference
is performed on an Anglican query, the query is run by a
particular inference algorithm. Inference algorithms must
provide an implementation for \texttt{infer}, as well as
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

Finally, Anglican queries use `primitive', or commonly known and
used, distributions, to draw random samples and condition
observations. Many primitive distributions are provided by
\iftoggle{full}{the \texttt{anglican.runtime} namespace}{the
runtime}, and an additional distribution can be defined by the
user by implementing a particular set of functions for the
distribution (via Clojure's protocol mechanism). The
\texttt{defdist} macro provides a convenient syntax for defining
primitive distributions.

@@

Having defined the query, we are now ready to run the query
using an inference algorithm. Function
\texttt{doquery}\iftoggle{full}{ provided
by the \texttt{anglican.core} namespace}{} accepts the inference
algorithm, the query, and optional parameters, and returns a
lazy sequence of samples. We use here the inference algorithm
called Lightweight Metropolis-Hastings (LMH). The algorithm is
somewhat slow to converge but can be used with any Anglican
query, and should be robust enough for our simple problem.

@@

\begin{figure}
	\centering
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
    \caption{Arrival time distribution for a single customer.}
    \label{fig:dist-same}
\end{figure}

\begin{figure}
	\centering
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
    \caption{Arrival time distributions for two customers.}
    \label{fig:dist-diff}
\end{figure}

@@

The compilation relies on the Clojure \textit{macro} facility,
and implemented as a library of
\textit{functions}\iftoggle{full}{ in namespace
\texttt{anglican.trap}, which are}{} invoked by macros.  The CPS
transformation is organized in top-down manner.  The top-level
function is  \texttt{cps-of-expression}, which receives an
expression and a continuation, and returns the expression in the
CPS form. For example, the CPS transformation of constant
\texttt{1} with continuation \texttt{cont} thus takes the
following form:
\begin{lstlisting}[style=default]
=> (cps-of-expression 1 'cont)
(cont 1 $state)
\end{lstlisting}

@@

In addition to checkpoints, there are a few other special forms
---  \texttt{store}, \texttt{retrieve}, \texttt{mem} --- which
modify program state. \iftoggle{full}{These forms are translated into
expressions involving calls of functions from the
\texttt{anglican.state} namespace. }{}The \texttt{mem} form, which
implements memoization, deserves a more detailed explanation.

@@

In the implementation of Anglican, inference algorithms are
instantiations of the (ad-hoc) polymorphic function
\texttt{infer} declared as Clojure's multimethod\iftoggle{full}{
in the \texttt{anglican.{\linebreak[0]}inference}
namespace}{}. The function accepts an algorithm identifier,
a query --- the probabilistic program in which to perform the
inference, an initial value for the query, and optional
algorithm parameters.

@@

An execution of the query may result in the following
sequence of checkpoint invocations:
\begin{lstlisting}[style=default]
  (sample 'C1 ...)
  (sample 'C2 ...)
  (sample 'C2 ...)
  (sample 'C1 ...)
  (sample 'C1 ...)
  (sample 'C1 ...)
  (sample 'C2 ...)
  (sample 'C3 ...)
\end{lstlisting}
According to the addressing scheme, the addresses generated
for this invocations are
\begin{lstlisting}[style=default]
  [C1 0][C2 0][C2 1][C1 16][C1 17][C1 18][C2 16][C3 0]
\end{lstlisting}
If the program is run by a Metropolis-Hastings algorithm, then a small change usually takes place in the sequence of checkpoints with each invocation, and the new sequence may become
\begin{lstlisting}[style=default]
  (sample 'C1 ...)
  (sample 'C2 ...)
  (sample 'C1 ...)
  (sample 'C1 ...)
  (sample 'C2 ...)
  (sample 'C2 ...)
  (sample 'C3 ...)
\end{lstlisting}
The addresses for the new sequence are
\begin{lstlisting}[style=default]
  [C1 0][C2 0][C1 16][C1 17][C2 16][C2 17][C3 0]
\end{lstlisting}
Despite the change, the correspondence between checkpoints of
similar types occurring in similar positions (relative positions
in contiguous subsequences of a certain type) is preserved, and
drawn values can be reused efficiently:

@@

This way, each occurrence of a checkpoint has unique address,
but small disturbances --- removal or addition of a single or just
a few checkpoints --- are unlikely to derail the entire sequence.
The probability of derailment depends on the padding. The
padding can be safely, and without any impact on performance,
set to rather large numbers. However, rounding up to a multiple
of 16 proved to be appropriate for all practical purposes.
\iftoggle{full}{%

Function \texttt{checkpoint-id} in the
\texttt{anglican.inference} namespace automates generation of
checkpoint addresses and can be used from any inference
algorithm.
}{%
}%
